---
title: "Homework12_Mohamud_N"
output: html_document
---
# Loading Libraries
```{r}
packs <- c("tidyverse", "spData", "terra", "tidyterra", "remotes", "sf", "gstat", "gridExtra", "cowplot", "caret", "ggplot2", "car")
lapply(packs, require, character.only = TRUE)


```

```{r loading datasets}
# 2018
respondent_18 <- read.csv("/Users/nimomohamud/Desktop/DS/2018/atusresp-2018/atusresp_2018.dat")
activity_18 <- read.csv("/Users/nimomohamud/Desktop/DS/2018/atussum-2018/atussum_2018.dat")
roster_18 <- read.csv("/Users/nimomohamud/Desktop/DS/2018/atusrost-2018/atusrost_2018.dat")
cps_18 <- read.csv("/Users/nimomohamud/Desktop/DS/2018/atuscps-2018/atuscps_2018.dat")
elder_18<- read.csv("/Users/nimomohamud/Desktop/DS/2018/atusrostec-2018/atusrostec_2018.dat")


# 2019
respondent_19 <- read.csv("/Users/nimomohamud/Desktop/DS/2019/atusresp-2019/atusresp_2019.dat")
activity_19 <- read.csv("/Users/nimomohamud/Desktop/DS/2019/atussum-2019/atussum_2019.dat")
roster_19 <- read.csv("/Users/nimomohamud/Desktop/DS/2019/atusrost-2019/atusrost_2019.dat")
cps_19 <- read.csv("/Users/nimomohamud/Desktop/DS/2019/atuscps-2019/atuscps_2019.dat")
elder_19<- read.csv("/Users/nimomohamud/Desktop/DS/2019/atusrostec-2019/atusrostec_2019.dat")


# 2020
respondent_20 <- read.csv("/Users/nimomohamud/Desktop/DS/2020/atusresp-2020/atusresp_2020.dat")
activity_20 <- read.csv("/Users/nimomohamud/Desktop/DS/2020/atussum-2020/atussum_2020.dat")
roster_20 <- read.csv("/Users/nimomohamud/Desktop/DS/2020/atusrost-2020/atusrost_2020.dat")
cps_20 <- read.csv("/Users/nimomohamud/Desktop/DS/2020/atuscps-2020/atuscps_2020.dat")
elder_20<- read.csv("/Users/nimomohamud/Desktop/DS/2020/atusrostec-2020/atusrostec_2020.dat")


# 2021
respondent_21 <- read.csv("/Users/nimomohamud/Desktop/DS/2021/atusresp-2021/atusresp_2021.dat")
activity_21 <- read.csv("/Users/nimomohamud/Desktop/DS/2021/atussum-2021/atussum_2021.dat")
roster_21 <- read.csv("/Users/nimomohamud/Desktop/DS/2021/atusrost-2021/atusrost_2021.dat")
cps_21 <- read.csv("/Users/nimomohamud/Desktop/DS/2021/atuscps-2021/atuscps_2021.dat")
elder_21<- read.csv("/Users/nimomohamud/Desktop/DS/2021/atusrostec-2021/atusrostec_2021.dat")


```

```{r merging file by year}
# Merge files for 2018

data_18 <- respondent_18 %>%
  inner_join(activity_18, by = "TUCASEID", suffix = c(".respondent", ".activity")) %>%
  inner_join(roster_18, by = "TUCASEID", suffix = c("", ".roster")) %>%
  inner_join(cps_18, by = "TUCASEID", suffix = c("", ".cps")) %>%
  inner_join(elder_18, by = "TUCASEID", suffix = c("", ".elder"))


# Merge files for 2019
data_19 <- respondent_19 %>%
  inner_join(activity_19, by = "TUCASEID", suffix = c(".respondent", ".activity")) %>%
  inner_join(roster_19, by = "TUCASEID", suffix = c("", ".roster")) %>%
  inner_join(cps_19, by = "TUCASEID", suffix = c("", ".cps"))%>%
  inner_join(elder_19, by = "TUCASEID", suffix = c("", ".elder"))


# Merge files for 2020
data_20 <- respondent_20 %>%
  inner_join(activity_20, by = "TUCASEID", suffix = c(".respondent", ".activity")) %>%
  inner_join(roster_20, by = "TUCASEID", suffix = c("", ".roster")) %>%
  inner_join(cps_20, by = "TUCASEID", suffix = c("", ".cps"))%>%
  inner_join(elder_20, by = "TUCASEID", suffix = c("", ".elder"))

# Merge files for 2021
data_21 <- respondent_21 %>%
  inner_join(activity_21, by = "TUCASEID", suffix = c(".respondent", ".activity")) %>%
  inner_join(roster_21, by = "TUCASEID", suffix = c("", ".roster")) %>%
  inner_join(cps_21, by = "TUCASEID", suffix = c("", ".cps"))%>%
  inner_join(elder_21, by = "TUCASEID", suffix = c("", ".elder"))


```

```{r combinding data}
# finding common cols across all datasets
common_cols <- intersect(intersect(names(data_18), names(data_19)), 
                         intersect(names(data_20), names(data_21)))

# selecting only the common cols from each data set
same_c_data_18 <- dplyr::select(data_18, all_of(common_cols))
same_c_data_19 <- dplyr::select(data_19, all_of(common_cols))
same_c_data_20 <- dplyr::select(data_20, all_of(common_cols))
same_c_data_21 <- dplyr::select(data_21, all_of(common_cols))

# binding the rows to prevent is.n
data_pre_pandemic <- bind_rows(same_c_data_18, same_c_data_19)
data_pandemic <- bind_rows(same_c_data_20, same_c_data_21)


combined_data <- bind_rows(data_pre_pandemic %>% mutate(period = "Pre-Pandemic"),data_pandemic %>% mutate(period = "Pandemic"))

# combine TRTEC from 2018-2021 with corresponding year and period labels

combined_trtec <- bind_rows(
  data_18 %>% select(TRTEC = TRTEC.activity) %>% mutate(year = 2018, period = "Pre-Pandemic"),
  data_19 %>% select(TRTEC = TRTEC.activity) %>% mutate(year = 2019, period = "Pre-Pandemic"),
  data_20 %>% select(TRTEC = TRTEC.activity) %>% mutate(year = 2020, period = "Pandemic"),
  data_21 %>% select(TRTEC = TRTEC.activity) %>% mutate(year = 2021, period = "Pandemic")
)

#Weekend/weekday vars: 
combined_data <- combined_data %>%
  mutate(
    day_type = case_when(
      TUDIARYDAY.respondent %in% c(1, 7) ~ "Weekend",  # 1 = Sunday, 7 = Saturday
      TUDIARYDAY.respondent %in% 2:6 ~ "Weekday"       # 2-6 = Monday to Friday
    )
  )

#month and season vars 
combined_data <- combined_data %>%
  mutate(
    #getting the month from the date format 
    Month = substr(as.character(TUDIARYDATE), 5, 6),
    season = case_when(
      Month %in% c("12", "01", "02") ~ "Winter",
      Month %in% c("03", "04", "05") ~ "Spring",
      Month %in% c("06", "07", "08") ~ "Summer",
      Month %in% c("09", "10", "11") ~ "Fall"
    )
  )


```

#Childcare 
## analysis 

```{r chilcare analysis }
# Childcare Analysis

# Adding childcare variables to combined data
combined_childcare <- bind_rows(
  same_c_data_18 %>% select(TUCASEID, TRTCC, TRTHH = TRTHH.respondent, TRTOHHCHILD, TRNUMHOU, TEAGE, TRSPPRES = TRSPPRES.respondent, TRNHHCHILD, TROHHCHILD, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2018, period = "Pre-Pandemic"),
  same_c_data_19 %>% select(TUCASEID, TRTCC, TRTHH = TRTHH.respondent, TRTOHHCHILD, TRNUMHOU, TEAGE, TRSPPRES = TRSPPRES.respondent, TRNHHCHILD, TROHHCHILD, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2019, period = "Pre-Pandemic"),
  same_c_data_20 %>% select(TUCASEID, TRTCC, TRTHH = TRTHH.respondent, TRTOHHCHILD, TRNUMHOU, TEAGE, TRSPPRES = TRSPPRES.respondent, TRNHHCHILD, TROHHCHILD, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2020, period = "Pandemic"),
  same_c_data_21 %>% select(TUCASEID, TRTCC, TRTHH = TRTHH.respondent, TRTOHHCHILD, TRNUMHOU, TEAGE, TRSPPRES = TRSPPRES.respondent, TRNHHCHILD, TROHHCHILD, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2021, period = "Pandemic")
)

# Adding weekend/weekday, month, and season variables
combined_childcare <- combined_childcare %>%
  mutate(
    day_type = case_when(
      TUDIARYDAY %in% c(1, 7) ~ "Weekend",  # 1 = Sunday, 7 = Saturday
      TUDIARYDAY %in% 2:6 ~ "Weekday"       # 2-6 = Monday to Friday
    ),
    Month = substr(as.character(TUDIARYDATE), 5, 6),  # Extract month from YYYYMMDD format
    season = case_when(
      Month %in% c("12", "01", "02") ~ "Winter",
      Month %in% c("03", "04", "05") ~ "Spring",
      Month %in% c("06", "07", "08") ~ "Summer",
      Month %in% c("09", "10", "11") ~ "Fall"
    )
  )

# Creating subsets for single- and multi-adult households
# Fix child count logic
single_adult_child <- combined_childcare %>%
  filter(
    TRNUMHOU >= 2,                 # At least 2 people in the household
    TRSPPRES == 3,                 #No spouse or unmarried partner present
    TEAGE > 17,                    # Respondent is an adult
    (TRNUMHOU - 1) * 13 > TEAGE   # Remaining members must be under 13
  ) %>%
  mutate(household_type = "Single-Adult")

multi_adult_child <- combined_childcare %>%
  filter(
    TRNUMHOU >= 3,                       # At least 3 people in the household (2 adults + children)
    TRSPPRES %in% c(1, 2),               # Spouse or unmarried partner present
    TEAGE > 17,                          # Respondent is an adult
    (TRNUMHOU - 2) * 13 > TEAGE         # Remaining members must all be under 13
  ) %>%
  mutate(household_type = "Multi-Adult")






# Combine datasets for plotting
child_combined_householdstypes <- bind_rows(single_adult_child, multi_adult_child)
sum(is.na(child_combined_householdstypes))

summary(child_combined_householdstypes$TRTCC)
```


## visuallization 
#Did the time spent on childcare activities differ between single-adult and multi-adult households during the pandemic years (2020-2021) compared to pre-pandemic years (2018-2019)?
```{r Pandemic Impact Question}
# Violin plot: Childcare time by household type and period
child_combined_householdstypes$period <- factor(child_combined_householdstypes$period, 
                                                levels = c("Pre-Pandemic", "Pandemic"))

ggplot(child_combined_householdstypes, aes(x = period, y = TRTCC, fill = household_type)) +
  geom_violin(trim = FALSE, alpha = 0.7) +  # Add violin plot
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA) +  # Add box plots
  labs(title = "Violin and Box Plot of Childcare Time by Period and Household Type",
       x = "Period",
       y = "Childcare Time",
       fill = "Household Type") +
  theme_minimal()
ggplot(child_combined_householdstypes, aes(x = factor(year), y = TRTCC, fill = household_type)) +
  geom_violin(trim = FALSE, alpha = 0.7, position = position_dodge(0.9)) +  # Add violin plot with dodge
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA, alpha = 0.9) +  # Add box plot with proper dodge
  labs(title = "Violin and Box Plot of Childcare Time by Year and Household Type",
       x = "Year",
       y = "Childcare Time",
       fill = "Household Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability



# Density plot: Childcare time by period and household type
ggplot(child_combined_householdstypes, aes(x = TRTCC, fill = household_type)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ period) +
  labs(title = "Density Plot of Childcare Time by Period and Household Type",
       x = "Childcare Time",
       y = "Density",
       fill = "Household Type") +
  theme_minimal()

```
#Did childcare time vary significantly across seasons (Winter, Spring, Summer, Fall)? Was this variation different for single-adult vs. multi-adult households?

```{r Season Trends Questions}
# Violin plot: Childcare time by season and household type
ggplot(child_combined_householdstypes, aes(x = season, y = TRTCC, fill = household_type)) +
  geom_violin(trim = FALSE, position = position_dodge(0.9), alpha = 0.7) +  
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA) +  
  facet_wrap(~ period) +  # Facet by period
  labs(title = "Violin and Box Plot of Childcare Time by Year, Period, and Household Type",
       x = "Year",
       y = "Childcare Time",
       fill = "Household Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

# Density plot: Childcare time by season
ggplot(child_combined_householdstypes, aes(x = TRTCC, fill = season)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ household_type) +
  labs(title = "Density Plot of Childcare Time by Season and Household Type",
       x = "Childcare Time",
       y = "Density",
       fill = "Season") +
  theme_minimal()
```
#Is there a difference in childcare time between weekdays and weekends for single-adult and multi-adult households? Did these patterns change during the pandemic?

```{r Days of the Week Question}

ggplot(child_combined_householdstypes, aes(x = day_type, y = TRTCC, fill = household_type)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  facet_wrap(~ period) +
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA, alpha = 0.9) +
  labs(title = "Childcare Time by Day of the Week and Household Type",
       x = "Day of the Week",
       y = "Childcare Time",
       fill = "Household Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 


# Violin plot: Childcare time by day type and period
ggplot(child_combined_householdstypes, aes(x = day_type, y = TRTCC, fill = household_type)) +
  geom_violin(trim = FALSE) +
  facet_wrap(~ year) +
  labs(title = "Violin Plot of Childcare Time by Day Type, Period, and Household Type",
       x = "Day Type",
       y = " Childcare Time",
       fill = "Household Type") +
  theme_minimal()

# Density plot: Childcare time by day type
ggplot(child_combined_householdstypes, aes(x = TRTCC, fill = day_type)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ household_type) +
  labs(title = "Density Plot of Childcare Time by Day Type and Household Type",
       x = " Childcare Time",
       y = "Density",
       fill = "Day Type") +
  theme_minimal()
```


#model selection
```{r all subset }
library(leaps)

####all- subset: better choice due to the small predicetor size
library(leaps)
all_subsets <- regsubsets(TRTCC ~ household_type * period +
                            household_type * day_type +
                            season * period +
                            day_type * period,
                          data = child_combined_householdstypes,
                          nbest = 1,  # it only keeps the best model of each size
                          nvmax = NULL,  # we are asking it to explore all subset sizes
                          method = "exhaustive")
summary(all_subsets)
#best model based on BIC
best_model <- summary(all_subsets)$which[which.min(summary(all_subsets)$bic), ]
child_best_model <- lm(
  TRTCC ~ household_type * period +
    household_type * day_type +
    period * season +
    period * day_type +
    season,
  data = child_combined_householdstypes
)

# Summary of the final model
summary(child_best_model)

#household_typeSingle-Adult:periodPandemic and seasonSummer are insig 
#partial f-test 
reduced_child_model <- lm(TRTCC ~ household_type * day_type + period * season + period * day_type + season, 
                    data = child_combined_householdstypes) 

anova(reduced_child_model, child_best_model) #The reduced model  explains the data almost as well as the full model 
summary(reduced_child_model)
plot(child_best_model)
summary(child_best_model)$r.squared


# Plot residuals vs. fitted values
plot(resid(reduced_child_model) ~ fitted(reduced_child_model),
     xlab = "Predicted Y", 
     ylab = "Residuals", 
     main = "Model Residuals",
     pch = 16, col = "blue")  # Add points for better visualization

# Add a horizontal line at 0 for reference
abline(h = 0, col = "red", lwd = 2)






```


```{r CV after model finalizaion }
library(caret)
trainIndex <- createDataPartition(child_combined_householdstypes$TRTCC, p = .8, list = FALSE)
trainData <- child_combined_householdstypes[trainIndex, ]
testData <- child_combined_householdstypes[-trainIndex, ]
best_model_cv <- lm(formula(child_best_model), data = trainData)
pred <- predict(best_model_cv, newdata = testData)
mean((testData$TRTCC - pred)^2)  # Mean Squared Error 69583.78



```

```{r mc}
library(usdm)
library(car)
predictors <- child_combined_householdstypes[, c("household_type", "period", "day_type", "season")]
predictors <- data.frame(lapply(predictors, function(x) {
  if (is.factor(x) || is.character(x)) as.numeric(as.factor(x)) else x
}))
vif_results <- vifstep(predictors, th = 10)  # Adjust the threshold if needed
print(vif_results)


#No variable from the 4 input variables has collinearity problem. 
```


```{r random forest}
library(randomForest)
library(caret)

rf_model <- randomForest(TRTCC ~ household_type + period + day_type + season, data = child_combined_householdstypes)
print(rf_model)

importance(rf_model)
varImpPlot(rf_model)

# k 10-fold cv 
cv_control <- trainControl(
  method = "cv",        
  number = 10,            
  verboseIter = TRUE      
)

# Train the Random Forest model with cross-validation
rf_cv_model <- train(
  TRTCC ~ household_type + period + day_type + season,  
  data = child_combined_householdstypes,               
  method = "rf",    # Random Forest method
  trControl = cv_control,      # cv control that i defined earlier 
  tuneGrid = expand.grid(.mtry = c( 3))   
)

# Print the results of cross-validation
print(rf_cv_model)

# Best model summary
rf_cv_model$bestTune  # Show the best hyperparameter (mtry)

varImpPlot(rf_model)

pred <- predict(rf_cv_model, newdata = child_combined_householdstypes)
plot(child_combined_householdstypes$TRTCC, pred,
     xlab = "Observed TRTCC", ylab = "Predicted TRTCC",
     main = "Observed vs. Predicted TRTCC")
abline(0, 1, col = "red", lwd = 2)


```

```{r comparison table}
# Load necessary libraries
library(broom)

lm_equation <- "TRTCC ~ household_type * day_type + period * season + period * day_type + season"
rf_equation <- "TRTCC ~ household_type + period + day_type + season"
# grabbing all the metrics from the lm
lm_adj_r2 <- summary(reduced_child_model)$adj.r.squared  # Adjusted R²
lm_aic <- AIC(reduced_child_model)                      # AIC
lm_bic <- BIC(reduced_child_model)                      # BIC
lm_rmse <- sqrt(mean(resid(reduced_child_model)^2))     # RMSE (calculated manually)

# grabbing all the metrics from the rfm
rf_r2 <- rf_cv_model$results$Rsquared[which.min(rf_cv_model$results$RMSE)]  # R² from cross-validation
rf_rmse <- min(rf_cv_model$results$RMSE)                                   # Best RMSE
rf_mae <- rf_cv_model$results$MAE[which.min(rf_cv_model$results$RMSE)]     # MAE

#creating comparison table
model_comparison <- data.frame(
  Model = c("Linear Model (LM)", "Random Forest (RF)"),
  Equation = c(lm_equation, rf_equation),
  `Adjusted R²` = c(lm_adj_r2, rf_r2),
  AIC = c(lm_aic, NA),  #aic for lm but unavailable for the rfm                          
  BIC = c(lm_bic, NA), #BIC for lm but unavailable for the rfm   
  RMSE = c(lm_rmse, rf_rmse), #both have RMSE
  MAE = c(NA, rf_mae), # MAE for rfm but unabilable for lm
  `% Variance Explained` = c(NA, rf_r2 * 100)  # % Variance Explained only avialibale for rfm
)
print(model_comparison)


```




```{r decision tree}
library(rpart)
library(rpart.plot)
tree_model <- rpart(
  TRTCC ~ household_type + period + day_type + season,
  data = child_combined_householdstypes,
  method = "poisson")

rpart.plot(tree_model)



```

```{r dt}

library(rpart)
tree_model <- rpart(household_type + period + day_type + season, 
                    data = child_combined_householdstypes)
plot(tree_model)
text(tree_model)


```

#Eldercare 

# Eldercare Analysis
```{r combined eldercare data}
# Adding eldercare variables to combined data
combined_eldercare <- bind_rows(
  same_c_data_18 %>% select(TUCASEID, TRTEC = TRTEC.respondent, TRELHH, TEELDUR, TRNUMHOU, TEAGE, TEAGE_EC, TRSPPRES = TRSPPRES.respondent, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2018, period = "Pre-Pandemic"),
  same_c_data_19 %>% select(TUCASEID, TRTEC = TRTEC.respondent, TRELHH, TEELDUR, TRNUMHOU, TEAGE, TEAGE_EC, TRSPPRES = TRSPPRES.respondent, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2019, period = "Pre-Pandemic"),
  same_c_data_20 %>% select(TUCASEID, TRTEC = TRTEC.respondent, TRELHH, TEELDUR, TRNUMHOU, TEAGE, TEAGE_EC, TRSPPRES = TRSPPRES.respondent, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2020, period = "Pandemic"),
  same_c_data_21 %>% select(TUCASEID, TRTEC = TRTEC.respondent, TRELHH, TEELDUR, TRNUMHOU, TEAGE, TEAGE_EC, TRSPPRES = TRSPPRES.respondent, TUDIARYDAY = TUDIARYDAY.respondent, TUDIARYDATE) %>% mutate(year = 2021, period = "Pandemic")
)

# Adding day_type, month, and season columns
combined_eldercare <- combined_eldercare %>%
  mutate(
    day_type = case_when(
      TUDIARYDAY %in% c(1, 7) ~ "Weekend",
      TUDIARYDAY %in% 2:6 ~ "Weekday"
    ),
    Month = substr(as.character(TUDIARYDATE), 5, 6),
    season = case_when(
      Month %in% c("12", "01", "02") ~ "Winter",
      Month %in% c("03", "04", "05") ~ "Spring",
      Month %in% c("06", "07", "08") ~ "Summer",
      Month %in% c("09", "10", "11") ~ "Fall"
    )
  )

# Creating subsets for single- and multi-adult households
single_adult_eldercare <- combined_eldercare %>%
  filter(
    TRNUMHOU >= 2,                 # Exactly two people in the household (one adult + one elder)
    TEAGE > 17,                    # Respondent is an adult
    TRSPPRES == 3,                 # No spouse or unmarried partner present
    TRTEC >= 0 # there are 42,634/ 2,323,231 have -1 values 
  ) %>%
  mutate(household_type = "Single-Adult")

multi_adult_eldercare <- combined_eldercare %>%
  filter(
    TRNUMHOU >= 3,                 # At least 3 people in the household
    TRSPPRES %in% c(1, 2),         # Spouse or unmarried partner present
    TEAGE > 17,                    # Respondent is an adult
    TRTEC >= 0 # there are 42,634/ 2,323,231 have -1 values 
  ) %>%
  mutate(household_type = "Multi-Adult")

# Combine both datasets for plotting
elder_combined_householdstypes <- bind_rows(single_adult_eldercare, multi_adult_eldercare) 
sum(is.na(elder_combined_householdstypes))


```
#Did the time spent on eldercare activities differ between single-adult and multi-adult households during the pandemic years (2020-2021) compared to pre-pandemic years (2018-2019)?
```{r Pandemic Impact Question}
# Violin plot: eldercare time by household type and period
elder_combined_householdstypes$period <- factor(elder_combined_householdstypes$period, 
                                                levels = c("Pre-Pandemic", "Pandemic"))

ggplot(elder_combined_householdstypes, aes(x = period, y = TRTEC, fill = household_type)) +
  geom_violin(trim = FALSE, alpha = 0.7) +  # Add violin plot
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA) +  # Add box plots
  labs(title = "Violin and Box Plot of Eldercare Time by Period and Household Type",
       x = "Period",
       y = "Eldercare Time",
       fill = "Household Type") +
  theme_minimal()

```

#Did childcare time vary significantly across seasons (Winter, Spring, Summer, Fall)? Was this variation different for single-adult vs. multi-adult households?

```{r Season Trends Questions}
# Violin plot: Childcare time by season and household type
ggplot(elder_combined_householdstypes, aes(x = season, y = TRTEC, fill = household_type)) +
  geom_violin(trim = FALSE, position = position_dodge(0.9), alpha = 0.7) +  
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA) +  
  facet_wrap(~ period) +  # Facet by period
  labs(title = "Violin and Box Plot of Eldercare Time by Year, Period, and Household Type",
       x = "Year",
       y = "Eldercare Time",
       fill = "Household Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

# Density plot: Childcare time by season
ggplot(elder_combined_householdstypes, aes(x = TRTEC, fill = season)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ household_type) +
  labs(title = "Density Plot of Eldercare Time by Season and Household Type",
       x = "Eldercare Time",
       y = "Density",
       fill = "Season") +
  theme_minimal()

```

#Is there a difference in eldercare time between weekdays and weekends for single-adult and multi-adult households? Did these patterns change during the pandemic?

```{r Days of the Week Question}

ggplot(elder_combined_householdstypes, aes(x = day_type, y = TRTEC, fill = household_type)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  facet_wrap(~ period) +
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA, alpha = 0.9) +
  labs(title = "Eldercare Time by Day of the Week and Household Type",
       x = "Day of the Week",
       y = "Eldercare Time",
       fill = "Household Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 


# Violin plot: eldercare time by day type and period
ggplot(elder_combined_householdstypes, aes(x = day_type, y = TRTEC, fill = household_type)) +
  geom_violin(trim = FALSE) +
  facet_wrap(~ year) +
  geom_boxplot(width = 0.2, position = position_dodge(0.9), outlier.shape = NA, alpha = 0.9) +
  labs(title = "Violin Plot of Eldercare Time by Day Type, Period, and Household Type",
       x = "Day Type",
       y = " Eldercare Time",
       fill = "Household Type") +
  theme_minimal()

# Density plot: Childcare time by day type
ggplot(elder_combined_householdstypes, aes(x = TRTEC, fill = day_type)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ household_type) +
  labs(title = "Density Plot of Eldercare Time by Day Type and Household Type",
       x = " Eldercare Time",
       y = "Density",
       fill = "Day Type") +
  theme_minimal()
```
#elder model selection
```{r all subset }
library(leaps)

####all- subset: better choice due to the small predicetor size
library(leaps)
all_subsets2 <- regsubsets(TRTEC ~ household_type * period +
                            household_type * day_type +
                            season * period +
                            day_type * period,
                          data = elder_combined_householdstypes,
                          nbest = 1,  # it only keeps the best model of each size
                          nvmax = NULL,  # we are asking it to explore all subset sizes
                          method = "exhaustive")
summary(all_subsets2)
#best model based on BIC
best_model2 <- summary(all_subsets2)$which[which.min(summary(all_subsets2)$bic), ]
elder_best_model <- lm(TRTEC ~ household_type * period + day_type + season + 
                         period:season + period:day_type + household_type:period, 
                       data = elder_combined_householdstypes)


# Summary of the final model
summary(elder_best_model)



# Plot residuals vs. fitted values
plot(resid(elder_best_model) ~ fitted(elder_best_model),
     xlab = "Predicted Y", 
     ylab = "Residuals", 
     main = "Model Residuals",
     pch = 16, col = "blue")  # Add points for better visualization

# Add a horizontal line at 0 for reference
abline(h = 0, col = "red", lwd = 2)






```

```{r CV after model finalizaion }
library(caret)
trainIndex2 <- createDataPartition(elder_combined_householdstypes$TRTEC, p = .8, list = FALSE)
trainData2 <- elder_combined_householdstypes[trainIndex2, ]
testData2 <- elder_combined_householdstypes[-trainIndex2, ]
best_model_cv2 <- lm(formula(elder_best_model), data = trainData2)
pred2 <- predict(best_model_cv2, newdata = testData2)
mean((testData2$TRTEC - pred2)^2)  # Mean Squared Error 59737.71
```
```{r mc}
library(usdm)
library(car)
predictors <- child_combined_householdstypes[, c("household_type", "period", "day_type", "season")]
predictors <- data.frame(lapply(predictors, function(x) {
  if (is.factor(x) || is.character(x)) as.numeric(as.factor(x)) else x
}))
vif_results <- vifstep(predictors, th = 10)  # Adjust the threshold if needed
print(vif_results)


#No variable from the 4 input variables has collinearity problem. 
```


```{r random forest}
library(randomForest)
library(caret)

rf_model2 <- randomForest(TRTEC ~ household_type * period + day_type + season + 
                         period:season + period:day_type + household_type:period, data = elder_combined_householdstypes)
print(rf_model2)

importance(rf_model2)
varImpPlot(rf_model2)

# k 10-fold cv 
cv_control2 <- trainControl(
  method = "cv",        
  number = 10,            
  verboseIter = TRUE      
)

# Train the Random Forest model with cross-validation
rf_cv_model2 <- train(
  TRTEC ~ household_type * period + day_type + season + 
                         period:season + period:day_type + household_type:period,  
  data = elder_combined_householdstypes,               
  method = "rf",    # Random Forest method
  trControl = cv_control,      # cv control that i defined earlier 
  tuneGrid = expand.grid(.mtry = c(1,2, 3,4,5))   
)

# Print the results of cross-validation
print(rf_cv_model2)

# Best model summary
rf_cv_model2$bestTune  # Show the best hyperparameter (mtry)

varImpPlot(rf_model2)

pred2 <- predict(rf_cv_model2, newdata = elder_combined_householdstypes)
plot(elder_combined_householdstypes$TRTEC, pred2,
     xlab = "Observed TRTEC", ylab = "Predicted TRTEC",
     main = "Observed vs. Predicted TRTEC")
abline(0, 1, col = "red", lwd = 2)


```

```{r comparison table}
# Load necessary libraries
library(broom)

lm_equation_elder <- "TRTEC ~ household_type * period + day_type + season + period:season + period:day_type + household_type:period"
rf_equation_elder <- "TRTEC ~ household_type * period + day_type + season + period:season + period:day_type + household_type:period"

# Grabbing all the metrics from the LM
lm_adj_r2_elder <- summary(elder_best_model)$adj.r.squared  
lm_aic_elder <- AIC(elder_best_model)                     
lm_bic_elder <- BIC(elder_best_model)                   
lm_rmse_cv_elder <- sqrt(mean((testData2$TRTEC - pred2)^2))  # Cross-validated RMSE  

# Grabbing all the metrics from the Random Forest model
rf_r2_elder <- rf_cv_model2$results$Rsquared[which.min(rf_cv_model2$results$RMSE)]  
rf_rmse_elder <- min(rf_cv_model2$results$RMSE)                                 
rf_mae_elder <- rf_cv_model2$results$MAE[which.min(rf_cv_model2$results$RMSE)]

# Creating the comparison table
model_comparison_elder <- data.frame(
  Model = c("Linear Model (LM)", "Random Forest (RF)"),
  Equation = c(lm_equation_elder, rf_equation_elder),
  `Adjusted R²` = c(lm_adj_r2_elder, rf_r2_elder),
  AIC = c(lm_aic_elder, NA),  # AIC for LM but unavailable for RF                          
  BIC = c(lm_bic_elder, NA),  # BIC for LM but unavailable for RF   
  RMSE = c(lm_rmse_cv_elder, rf_rmse_elder), # Both have RMSE
  MAE = c(NA, rf_mae_elder), # MAE for RF but unavailable for LM
  `% Variance Explained` = c(NA, rf_r2_elder * 100)  # % Variance Explained only available for RF
)
print(model_comparison_elder)



```




```{r decision tree}
library(rpart)
library(rpart.plot)
tree_model <- rpart(
  TRTEC ~ household_type + period + day_type + season,
  data = elder_combined_householdstypes,
  method = "poisson")

rpart.plot(tree_model)



```

